### 2016-11-26 dissertation/normal.tex
dissertation v2_part normal complete

\section{Motivation}
\label{sec:normal_motiv}
From the preliminaries about FSMs in section \ref{sec:reach}, we learn that the
Moore machine does not rely on inputs for state transitions. 
As depicted in Figure \ref{fig:Moore}(a), a typical Moore machine implementation
consists of combinational logic component and register files, where
$r_0,\dots,r_k$ are present state (PS) variables 
standing for state inputs (SI), and $r_0',\dots,r_k'$ are next state (NS) variables standing for
state outputs (SO). Figure \ref{fig:Moore}(b) shows the state transition graph (STG) of 
a Moore machine with $k+1$ distinct states. We notice that it forms a simple chain,
with $k$ consecutive transitions the machine reaches final state $R_k$.

@@

In practice,
some arithmetic components are designed in sequential circuits similar to the structure in 
Figure \ref{fig:Moore}(a). Initially the operands are loaded into the registers, 
then the closed circuit executes without taking any additional information from outside,
and store the results in registers after $k$ clock cycles. Its behavior can be described using
STG in Figure \ref{fig:Moore}(b): state $R$ denotes the bits stored in registers. Concretely, $R_init$ is the initial
state (usually reset to all zeros), $R_1$ to $R_{k-1}$ are intermediate results stored as SO of current state and SI
for next state, and $R_k$ (or $R_{final}$) is the final result given by arithmetic circuits (and equals to the answer
to arithmetic function when circuit is working functional correctly).
This kind of design results in 
reusing a smaller combinational logic component such that the area cost is greatly optimized.
However, it also brings difficulties in verifying the the circuit functions.

@@

% \section{Normal Basis Multiplier over Galois Field}
% From algebraic view, a field is a space, and field elements are dots in the space. Those elements can be 
% represented with unique coordinates, which requires the pre-definition of a basis vector. In this
% section, we discuss a special basis called normal basis, as well as the advantages adopting it in 
% GF operations esp. multiplication.
% \subsection{Normal Basis}
% Given a Galois field (GF) $\Fkk$ is a finite field with  $2^k$ elements and characteristic equals to 2.
% Its elements can be written in polynomials of $\alpha$, when there is an irreducible polynomial $p(\alpha)$
% defined.
% 
% If we use a basis $\{1,\alpha,\alpha^2,\alpha^3,\dots,\alpha^{k-1}\}$, we can easily transform polynomial representations
% to binary bit-vector representations by recording the coefficients. For example,
% 
% \begin{table}[H]
% \centering
% \caption{Bit-vector, Exponential and Polynomial representation of
% elements in  ${\mathbb{F}}_{2^4} = {\mathbb{F}}_2[x]
% \pmod{x^4+x^3+1}$}
% \begin{tabular}{|c|c||c|c|} 
% \hline
% $a_3a_2a_1a_0$ & Polynomial     &$a_3a_2a_1a_0$ & Polynomial  \\
% \hline
% $0000$        & $0$           & $1000$  &$\alpha^3$\\
% \hline
% $0001$        & $1$           & $1001$  & $\alpha^3 + 1$\\
% \hline
% $0010$        &  $\alpha$       & $1010$ & $\alpha^3 + \alpha$  \\
% \hline
% $0011$        &  $\alpha + 1$   & $1011$ &  $\alpha^3+\alpha+1$\\
% \hline
% $0100$        &  $\alpha^2$     &  $1100$ &  $\alpha^3 + \alpha^2$\\
% \hline
% $0101$        & $\alpha^2 + 1$ & $1101$  & $\alpha^3+\alpha^2+1$\\
% \hline
% $0110$        &  $\alpha^2 + \alpha$ & $1110$ &  $\alpha^3+\alpha^2+\alpha$\\
% \hline
% $0111$        & $\alpha^2+\alpha+1$ & $1111$ & $\alpha^3+\alpha^2+\alpha+1$\\
% \hline
% \end{tabular}
% \label{table:booltogalois}  
% \end{table}
% 
% Basis $\{1,\alpha,\alpha^2,\alpha^3,\dots,\alpha^{k-1}\}$ is called {\bf standard basis} (StdB), which results in
% a straightforward representation for elements, and operations of elements such as addition and subtraction.
% The addition/subtraction of GF elements in StdB follows the rules of polynomial addition/subtraction
% where coefficients belong to $\mathbb F_2$. In other words, using the definition of {\it exclusive or} (XOR) in
% Boolean algebra, element $A$ add/subtract by element $B$ in StdB is defined as
% \begin{align}\label{eqn:StdB}
% A+B = A-B &= (a_0,a_1,\dots,a_{k-1})_{StdB} \xor (b_0,b_1,\dots,b_{k-1})_{StdB} \nonumber\\
% &=(a_0\oplus b_0, a_1\oplus b_1,\dots,a_{k-1}\oplus b_{k-1})_{StdB} 
% \end{align}
% 
% \subsection{Multiplication using Normal Basis}
% Besides addition/subtraction, multiplication is also very common in arithmetic circuit design.
% The multiplication of GF elements in $\Fkk$ in StdB follows the rule of polynomial multiplication.
% However, it will result in $O(k^2)$ bitwise operations. In other words, if we implement GF multiplication
% in bit-level logic circuit, it will contain $O(k^2)$ gates. When the datapath size $k$ is large,
% the area and delay of circuit will be costly.
% 
% In order to lower down the complexity of arithmetic circuit design, Massey and Omura \cite{MasseyOmura} % ref 7 in RH paper
% use a new basis to represent GF elements, which is called {\bf normal basis} (NB).
% A normal basis over $\Fkk$ is written in the form of
% \begin{equation*}
% N.B. ~~~ \N = \{\beta,\beta^2,\beta^4,\beta^8,\dots,\beta^{2^{k-1}}\}
% \end{equation*}
% Respectively, a field element in NB representation is actually
% \begin{align*}
% A &= (a_0,a_1,\dots,a_{k-1})_{NB} \\
%   &= a_0\beta+a_1\beta^2+\cdots+a_{k-1}\beta^{2^{k-1}} \\
%   &= \sum_{i=0}^{k-1} a_i\beta^{2^i}
% \end{align*}
% 
% According to the definition, a normal basis is a vector where the next entry is the square of the former one.
% We note that the vector is cyclic, i.e. $\beta^{2^k} = \beta$ due to {\it Fermat's little theorem}.
% {\bf Normal element} $\beta$ is an element from the field which is used to construct the normal basis,
% and can be represent as a power of primitive element $\alpha$: 
% \begin{equation*}
% \beta = \alpha^t, ~~~ 1\leq t<2^k
% \end{equation*}
% 
% The addition and subtraction of elements in NB representation are similar to equation \ref{eqn:StdB}.
% However, what makes NB powerful is its property when doing multiplications and exponentiations.
% The following lemmas and examples illustrate this fabulous property very well.
% \begin{Lemma}[Square of NB]
% \label{lem:squareNB}
% In $\Fkk$, equation 
% \begin{equation*}
% (a+b)^2 = a^2 + b^2
% \end{equation*}
% has been proved. According to the \textbf{binomial theorem}, it can be extended to
% \begin{align*}
% &(b_0\beta + b_1\beta^2 + b_2\beta^4 + \dots + b_{k-1}\beta^{2^{k-1}})^2 \\
% &= b_0^2\beta^2 + b_1^2\beta^4 + b_2^2\beta^8 + \dots + b_{k-1}^2\beta \\
% &= b_{k-1}^2\beta + b_0\beta^2 + b_1\beta^4 + \dots + b_{k-2}\beta^{2^{k-1}}
% \end{align*}
% \end{Lemma}
% This lemma concludes that the square of an element in NB equals to a simple right-cyclic shift of the bit-vector.
% Obviously, StdB representation does not have this benefit.
% 
% \begin{Example}[Square of NB]
% In GF $\mathbb F_{2^3}$ constructed by irreducible polynomial $x^3 + x + 1$, the standard basis is denoted as 
% $\{ 1, \alpha, \alpha^2\}$ where $\alpha^3+\alpha+1=0$.
% Let $\beta = \alpha^3$, then $\N = \{ \beta, \beta^2, \beta^4\}$ forms a normal basis. 
% Write down element $E$ using both representations:
% \begin{align*}
% E &= (a_0,a_1,a_2)_{StdB} = (b_0,b_1,b_2)_{NB} \\
%   &= a_0 + a_1\alpha + a_2\alpha^2 = b_0\beta + b_1\beta^2 + b_2\beta^4
% \end{align*}
% Compute the square of $E$ in StdB first:
% \begin{align*}
% E^2 &= a_0 + a_1\alpha^2 + a_2\alpha^4 \\
%     &= a_0 + a_2\alpha + (a_1 + a_2)\alpha^2 \\
%     &= (a_0,a_2,a_1+a_2)_{StdB}
% \end{align*}
% When it is computed in NB, we can make it very simple:
% \begin{align*}
% E^2 &= \overset{\xrightarrow{Cyclic~~shift}}{(b_0,b_1,b_2)}_{NB} \\
% 	&= (b_2,b_0,b_1)_{NB}
% \end{align*}
% \end{Example}
% 
% This example shows that convenience to use NB when computing $2^k$ power of an element.
% Multiplication is a bit complicated than squaring; but when it is decomposed as bit-wise
% operations, the property in lemma \ref{lem:squareNB} can be well utilized.
% 
% \begin{Example}[Bit-wise NB multiplication]
% Assume there are 2 binary vectors representing 2 operands in NB over $\Fkk$: 
% $A = (a_0, a_1, \dots, a_{k-1}), B = (b_0, b_1, \dots, b_{k-1})$. Note that in this example, 
% by default we use normal basis representation so subscript ``NB" is skipped. Their product can also be written
% as: $$C = A\times B = (c_0, c_1, \dots, c_{k-1})$$
% 
% Assume the most significant bit (MSB) of the product can be represented by a function $f_{mult}$: 
% \begin{equation}
% \label{eqn:multiMSB}
% c_{n-1} = f_{mult}(a_0, a_1, \dots, a_{n-1}; b_0, b_1, \dots, b_{n-1})
% \end{equation}
% Before discussing the details of the function $f_{mult}$, we can take 
% a square on both side of equation \ref{eqn:multiMSB}, i.e. $C^2 = A^2\times B^2$.
% Obviously, using the property in lemma \ref{lem:squareNB}, the original second most significant bit 
% becomes the new MSB because of right-cyclic shifting. 
% Concretely, 
% $$(c_{k-1},c_0,c_1,\dots,c_{k-2}) = (a_{k-1},a_0,a_1,\dots,a_{k-2})\times(b_{k-1},b_0,b_1,\dots,b_{k-2})$$
% Note $A^2, B^2$ and $C^2$ still belong to $\Fkk$, thus as a universal function implementing MSB multiplication
% over $\Fkk$, $f_{mult}$ still keeps the same. As a result, the new MSB can be written as 
% \begin{equation}
% \label{eqn:shiftMSB}
% c_{k-2} = f_{mult}(a_{k-1}, a_0, a_1, 
% \dots, a_{k-2}; b_{k-1}, b_0, b_1, \dots, b_{k-2})
% \end{equation}
% Similarly, if we take a square again on the new equation, we can get $c_{k-3}$.
% Successively we can derive all bits of product $C$ using the same function $f_{mult}$, and the only
% adjustment we need to make is to right-cyclic shift 2 operands by 1 bit each time.
% \end{Example}
% 
% From above example, it is proved that a universal structure that implements $f_{mult}$ can be reused
% for $k$ times in NB multiplication over $\Fkk$. Comparing to StdB, which requires distinct design 
% for every bit of multiplication, NB is less costly if we can prove $f_{mult}$ will not result in 
% a structure with $O(k^2)$ complexity. So our next mission is to explore the details of $f_{mult}$
% to prove it will be a relatively simple design with complexity lower than $O(k^2)$.
% 
% If we want to make the complexity of $f_{mult}$ lower than $O(k^2)$, then the best choice is to try out 
% linear functions. As we know, matrix multiplication can simulate all possible combinations of 
% linear functions (which is also the reason it is used as basic model to simulate the behavior of a neuron
% in neural network machine learning algorithms). Imagine $A$ is a $k$-bit row vector and $B$ is a $k$-bit 
% column vector, then the single bit product can be written as the product of matrix multiplication
% $$c_{l} = A\times C\times B$$
% where $C$ is a $k\times k$ square matrix.
% 
% \begin{Definition}[$\lambda$-Matrix]
% A binary $k\times k$ matrix $M$ is used to describe the bit-wise normal basis multiplication function $f_{mult}$ where
% \begin{equation}
% \label{eqn:def_lambda}
% c_{l} = f_{mult}(A, B) = A \times M \times B^T
% \end{equation}
% $B^T$ denotes vector transposition. Matrix $M$ is called $\lambda$-Matrix of
% $k$-bit NB multiplication over $\Fkk$.
% \end{Definition} 
% 
% When taking different bits $l$ of the product in equation \ref{eqn:def_lambda}, 
% we obtain a series of conjugate matrices of $M$. 
% Which means instead of shifting operands $A$ and $B$, we can also shift the 
% matrix.
% 
% More specifically, we denote the matrix by \emph{$l$-th $\lambda$-Matrix} as 
% $$c_l = A \times M^{(l)} \cdot B^T$$
% Meanwhile, the operator shifting rule in equation \ref{eqn:shiftMSB} still holds. Then we have relation 
% $$c_{l-1} = A \cdot M^{(l-1)} \cdot B^T = shift(A) \cdot M^{(l)} \cdot shift(B)^T$$ which means
% by right and down cyclically shifting $M^{(l-1)}$, we can get $M^{(l)}$.
% 
% \begin{Example}[NB multiplication using $\lambda$-Matrix]
% Over GF $\mathbb F_{2^3}$ constructed by irreducible polynomial $\alpha^3 + \alpha + 1$, let normal element $\beta = \alpha^3$, $N = \{ \beta, \beta^2, \beta^4\}$ 
% forms a normal basis. Corresponding $0$-th $\lambda$-Matrix is
% \begin{equation*}
% M^{(0)} = \left(
% \begin{array} {lcr}
% 0 & 1 & 0\\
% 1 & 0 & 1\\
% 0 & 1 & 1
% \end{array} \right).
% \end{equation*}
% i.e.,
% \begin{equation*}
% c_0 = (a_0\  a_1\  a_2)\left(
% \begin{array} {lcr}
% 0 & 1 & 0\\
% 1 & 0 & 1\\
% 0 & 1 & 1
% \end{array} \right)\left(
% \begin{array} {lcr}
% b_0\\
% b_1\\
% b_2
% \end{array} \right)
% \end{equation*}
% From $0$-th $\lambda$-Matrix we can directly write down all remaining $\lambda$-Matrices:
% \begin{equation*}
% M^{(1)} = \left(
% \begin{array} {lcr}
% 1 & 0 & 1\\
% 0 & 0 & 1\\
% 1 & 1 & 0
% \end{array} \right)~~~~~~
% M^{(2)} = \left(
% \begin{array} {lcr}
% 0 & 1 & 1\\
% 1 & 1 & 0\\
% 1 & 0 & 0
% \end{array} \right)
% \end{equation*}
% \end{Example}
% 
% If we generalize the definition and explore the nature of $\lambda$-Matrix, it is defined as cross-product terms from multiplication, which is 
% \begin{equation}
% Product~vector~C = (\sum_{i=0}^{k-1}a_i\beta^{2^i})(\sum_{j=0}^{k-1}b_j\beta^{2^j}) = \sum_{i=0}^{k-1}\sum_{j=0}^{k-1}a_ib_j\beta^{2^i}\beta^{2^j}
% \end{equation}
% The expressions $\beta^{2^i}\beta^{2^j}$ are referred to as cross-product terms, and can be represented by
% NB, i.e.
% \begin{equation}
% \beta^{2^i}\beta^{2^j} = \sum_{l=0}^{k-1}\lambda_{ij}^{(l)}\beta^{2^l}, \ \ \lambda_{ij}^{(l)} \in \mathbb F_2.
% \end{equation}
% Substitution yields, result is an expression for l-th digit of product as showed in equation \ref{eqn:multiMSB}:
% \begin{equation}
% c_l = \sum_{i=0}^{k-1}\sum_{j=0}^{k-1}\lambda_{ij}^{(l)}a_ib_j
% \end{equation}
% $\lambda_{ij}^{(l)}$ is the entry with coordinate $(i,j)$ in $l$-th $\lambda$-Matrix.
% 
% The $\lambda$-Matrix can be implemented with XOR and AND gates in circuit design.
% The very naive implementation requires $O(C_N)$ gates, where $C_N$ is the number of
% nonzero entries in $\lambda$-Matrix.
% There usually exists multiple NBs in $\Fkk, k>3$. If we employ a random NB, there is no mathematical
% guarantee that $C_N \sim o(k)$ (symbol $o$ denotes ``strictly lower than bound"). However,
% Mullin et al. \cite{mullinONB} % This citation is valid
% proves that in certain GF $\mathbb F_{p^{k_{opt}}}$, there always exists at least one NB such that 
% its corresponding $\lambda$-Matrix has $C_N = 2n-1$ nonzero entries. A basis with this property is
% called optimal normal basis (ONB), details are introduced in appendix \ref{append:ONB}.
% 
% In practice, large size NB multipliers are usually designed in $\Fkk$ when ONB exists
% to minimized the number of gates. So in the following part of this chapter and our experiments,
% we only focus on ONB multipliers instead of general NB multipliers.
% 
% 
% % After fixing this, add a whole-piece StdB vs NB example, list their cost as the conclusion
% \subsection{Comparison between Standard Basis and Normal Basis}
% At the end of this section, a detailed example is used to make a comparison between StdB multiplication
% and NB multiplication.
% \begin{Example}[Rijndael's finite field]
% Rijndael uses a characteristic 2 finite field with 256 elements, which can also be called the GF $\mathbb F_{2^8}$.
% Let us define the primitive element $\alpha$ using irreducible polynomial 
% $\alpha^8+\alpha^7+\alpha^6+\alpha^4+\alpha^2+\alpha+1$. Coincidently, $\alpha$ is also a normal element,
% i.e. $\beta = \alpha$ can construct a NB $\{\alpha,\alpha^2,\alpha^4,\alpha^8,\alpha^{16},\alpha^{32},\alpha^{64},
% \alpha^{128}\}$.
% 
% We pick a pair of elements from the Rijndael's field: $A=(0100~1011)_{StdB} = (4B)_{StdB},~B=(1100~1010)_{StdB} = (CA)_{StdB}$.
% First let us compute their product in StdB, the rule follows ordinary polynomial multiplication.
% 
% \begin{align*}
% A\cdot B &= (\alpha^6+\alpha^3+\alpha+1)(\alpha^7+\alpha^6+\alpha^3+\alpha)\\
% &= (\alpha^{13}+\alpha^{10}+\alpha^8+\alpha^7)+(\alpha^{12}+\alpha^9+\alpha^7+\alpha^6)+
% 	(\alpha^9+\alpha^6+\alpha^4+\alpha^3)\\
% 	&~~~+(\alpha^7+\alpha^4+\alpha^2+\alpha) \\
% &= \alpha^{13}+\alpha^{12}+\alpha^{10}+\alpha^8+\alpha^7+\alpha^3+\alpha^2+\alpha
% \end{align*}
% Note that this polynomial is not the final form of the product because it needs to be
% reduced modulo irreducible polynomial $\alpha^8+\alpha^7+\alpha^6+\alpha^4+\alpha^2+\alpha+1$.
% This can be done using base-2 long division. Note the dividend and divisor are written in pseudo Boolean
% vectors, not real Boolean vectors in any kind of bases.
% 
% % The folowing part is customized latex code mimicing binary long division
% % Note \divrule coordinates may not reflect the intuitive column # in tabular. Adjust by urself!!!
% 
% \vspace{1cm}
% 
% \newdimen\digitwidth
% \settowidth\digitwidth{0}
% \def~{\hspace{\digitwidth}}
% 
% 
% \def\divrule#1#2{%
% \noalign{\moveright#1\digitwidth%
% \vbox{\hrule width#2\digitwidth}}}
% 111010111\,\begin{tabular}[b]{@{}r@{}}
% 101001 \\ \hline
% \big)\begin{tabular}[t]{@{}l@{}}
% 11010110001110 \\
% 111010111 \\ \divrule{0}{14}
% ~~111101101 \\
% ~~111010111 \\ \divrule{2}{12}
% ~~~~~111010110 \\
% ~~~~~111010111 \\ \divrule{5}{9}
% ~~~~~~~~~~~~~1
% \end{tabular}
% \end{tabular}
% \vspace{0.5cm}
% 
% The final remainder is $1$, i.e. the product equals to 1 in StdB.
% 
% On the other hand, operands $A$ and $B$ can be written in NB as
% $$A = (0010~1001)_{NB},~~B = (0100~0010)_{NB}$$
% The $\lambda$-Matrix for $\mathbb F_{2}[x] \pmod{x^8+x^7+x^6+x^4+x^2+x+1}$
% is
% (Computation of $\lambda$-Matrix refers to appendix \ref{append:NB})
% \begin{equation*}
% M^{(0)} = \left(\begin{array}{lccccccr}
% 0 &0 &0 &0 &1 &0 &1 &1 \\
% 0 &0 &1 &1 &1 &1 &0 &0 \\
% 0 &1 &0 &0 &0 &0 &1 &0 \\
% 0 &1 &0 &0 &1 &1 &0 &1 \\
% 1 &1 &0 &1 &0 &1 &0 &0 \\
% 0 &1 &0 &1 &1 &0 &0 &1 \\
% 1 &0 &1 &0 &0 &0 &0 &0 \\
% 1 &0 &0 &1 &0 &1 &0 &1
% \end{array}\right)
% \end{equation*}
% Taking matrix multiplication $c_0 = A\times M^{(0)}\times B^T$,
% the result is $c_0 = 1$. Then by cyclic shifting $A$ and $B$
% (or shifting $M^{(0)}$, either is applicable), we can successively
% obtain other bits of product. The final answer is
% $$C = (0000~0001)_{NB}$$
% It is equivalent to the result in StdB.
% \end{Example}
% 
% From the intuition of humans, StdB multiplication is straightforward and easier to understand
% while NB is difficult to comprehend. However, if we implement both multiplications to 
% hardware multipliers, it will be clear which side a circuit designer prefers.
% 
% % May consider providing details?
% 
% Mastrovito multiplier and Montgomery multiplier are 2 common designs
% of GF multipliers using StdB. As a naive implementation of GF multiplication,
% Mastrovito multiplier uses most number of gates:
% $k^2$ AND gates plus $k^2-\Delta$ XOR gates \cite{Mastrovito}. Montgomery multiplier 
% applies lazy reduction techniques and results in a better latency performance, while the number of gates are about
% the same with Mastrovito multiplier:
% $k^2$ AND gates plus $k^2-k/2$ XOR gates \cite{Montgomery}. 
% Concretely, typical design of Mastrovito multiplier consists of 218 logic gates, while 
% Montgomery multiplier needs 198 gates. However, the NB multiplier reuses the $\lambda$-Matrix 
% logic, so this component will only need to be implemented for once. 
% Consider the definition of matrix multiplication, it needs $C_N$ AND gates to apply 
% bit-wise multiplication and $C_N-1$ XOR gates to sum the intermediate products up. The number of nonzero entries
% in the $\lambda$-Matrix can be counted: $C_N = 27$.
% As a result, the most naive NB multiplier design (or Massey-Omura multiplier \cite{MasseyOmura})
% contains 53 gates in total, which is a great saving in area cost comparing to StdB multipliers.
% 
% \section{Design a Normal Basis Multiplier on Gate Level}
% The NB multiplier design consumes much less gates than ordinary StdB multiplier design, even if 
% we use the most naive design. However, the modern NB multiplier design has been improved a lot from the 
% very first design model proposed by Massey and Omura in 1986 \cite{MasseyOmura}. In order to 
% test our approach on practical contemporary circuits, it is necessary to learn the mechanism and design 
% routine of several kinds of modern NB multipliers.
% \subsection{Sequential Multiplier with Parallel Outputs}
% The major benefit of NB multiplier origins from the sequential design. A straightforward design implementing 
% the cyclic-shift of operands and $\lambda$-Matrix logic component is the Massey-Omura multiplier.
% 
% \begin{figure}[H]
% \centering{
% \includegraphics[width=\textwidth]{newfig/MasseyOmura.pdf}
% \caption{A typical SMSO structure of Massey-Omura multiplier}
% \label{fig:MasseyOmura}}
% \end{figure}
% 
% Figure \ref{fig:MasseyOmura} shows the basic architecture of a Massey-Omura multiplier. The operands
% $A$ and $B$ are 2 arrays of flip-flops which allow 1-bit right-cyclic shift every clock cycle.
% The logic gates in the boxes implements the matrix multiplication with $\lambda$-Matrix $M^{(0)}$, 
% while each AND gate corresponds to term $a_ib_j$ and each XOR gate corresponds to addition $a_ib_j+a_{i'}b_{j'}$. 
% The XOR layer has only 1 output, giving out 1 bit of product $C$ every clock cycle.
% 
% The behavior of Massey-Omura multiplier can be concluded as: pre-load operands $A,B$ and reset $C$ to 0, after 
% executing for $k$ clock cycles, the data stored in flip-flop array $C$ is the product $A\times B$.
% We note that there is only one output giving 1 bit of the product each clock cycle, which matches the 
% definition of serial output to communication channel. Therefore this type of design is named as 
% sequential multiplier with serial output (SMSO).
% The SMSO architecture need $C_N$ AND gates and $C_N-1$ XOR gates, which equals to $2k-1$ AND gates and $2k-2$ XOR gates
% if it is designed using ONB. In fact, the number of gates can be reduced if the multiplication is 
% implemented using a conjugate of SMSO.
% 
% The gate-level logic boxes are implementing following function:
% \begin{equation}
% \label{eqn:SMPOterms}
% c_l = row_1(A\times M^{(l)}) \times B + row_2(A\times M^{(l)}) \times B + \cdots + row_{k}(A\times M^{(l)}) \times B
% \end{equation}
% It can be decomposed into $k$ terms. If we only compute one term for each $c_l,~0\leq l\leq k-1$ in one 
% clock cycle, make $k$ outputs and add them up using shift register after $k$ clock cycles, 
% it will generate the same result with SMSO. This kind of architecture is named as 
% sequential multiplier with parallel outputs (SMPO). The basic SMPO, as a conjugate of Massey-Omura multiplier,
% is invented by Agnew et al. \cite{agnew1991implementation}.
% 
% \begin{figure}[H]
% \centering{
% \includegraphics[width=\textwidth]{newfig/mySMPO.eps}
% \caption{5-bit Agnew's SMPO. Index $i$ satisfies $0<i<4$, indices $u,v$ are determined by column \# of nonzero entries in $i$-th row of $\lambda$-Matrix $M^{(0)}$, i.e. if entry $M_{ij}^{(0)}$ is a nonzero entry, $u$ or $v$ equals to $i+j \pmod 5$. Index $w = 2i\pmod 5$}
% \label{fig:SMPO}}
% \end{figure}
% 
% \begin{Example}[5-bit Agnew's SMPO]
% Given GF $\mathbb F_{2^5}$ and primitive element $\alpha$ defined by irreducible polynomial 
% $\alpha^5+\alpha^2+1=0$, normal element $\beta = \alpha^5$ constructs an ONB $\{\beta,\beta^2,\beta^4,\beta^8,\beta^{16}\}$.
% The $0$-th $\lambda$-Matrix for this ONB is
% \begin{equation*}
% M^{(0)} = \left(\begin{array}{lcccr}
% 0 &1 &0 &0 &0 \\
% 1 &0 &0 &1 &0 \\
% 0 &0 &0 &1 &1 \\
% 0 &1 &1 &0 &0 \\
% 0 &0 &1 &0 &1
% \end{array}\right)
% \end{equation*}
% Then a typical design of 5-bit Agnew's SMPO is depicted in figure \ref{fig:SMPO}.
% 
% The operands part of this circuit is the same with Massey-Omura multiplier. The differences are on 
% the matrix multiplication part, while it is implemented as separate logic blocks for 5 outputs,
% and the 5 blocks are connected in a shift register fashion. By analyzing the detailed function of 
% logic blocks, we can reveal the mechanism of Agnew's SMPO.
% 
% Suppose we implement $M^{(0)}$ as the logic block in SMSO. In the first clock cycle, the output is
% \begin{equation}
% \label{eqn:5bitSMPO_c0}
% c_0 = a_1b_0+(a_0+a_3)b_1 + (a_3+a_4)b_2 + (a_1+a_2)b_3 + (a_2+a_4)b_4
% \end{equation}
% Note it is written in the form of equation \ref{eqn:SMPOterms}. In next clock cycles we can obtain 
% remaining bits of the product, which can be written in following general form polynomial:
% \begin{align}
% c_i =& b_ia_{i+1} + b_{i+1}(a_i + a_{i+3}) + b_{i+2}(a_{i+3} + a_{i+4}) \nonumber\\
% &+ b_{i+3}(a_{i+1} + a_{i+2}) + b_{i+4}
% (a_{i+2} + a_{i+4}),\ 0\leq i\leq 4 \nonumber
% \end{align}
% Note all index calculations are reduced modulo 5.
% 
% Now let us observe the behavior of 5-bit Agnew's SMPO. Initially all DFFs are reset to 0. 
% In the first clock cycle,
% signal sent to the flip-flop in block $R_0$ denotes function:
% $$R_0^{(1)} = a_1b_0$$
% It equals to the first term of equation \ref{eqn:5bitSMPO_c0}. In the second clock cycle,
% this signal is sent to block $R_1$ through wire $r_0$, and this block also receives data 
% from operands (shifted by 1 bit), generating signal $a_u,a_v$ and $b_w$. Concretely,
% signal sent to flip-flop in block $R_1$ is:
% $$R_1^{(2)} = R_0^{(1)} + (a_0+a_3)b_1 = a_1b_0+(a_0+a_3)b_1$$
% which forms first 2 terms of equation \ref{eqn:5bitSMPO_c0}. Similarly, we track the signal 
% on $R_2$ in third clock cycle, signal on $R_3$ in fourth clock cycle, finally we can get 
% $$R_4^{(5)} = a_1b_0+(a_0+a_3)b_1 + (a_3+a_4)b_2 + (a_1+a_2)b_3 + (a_2+a_4)b_4$$
% which equals to $c_0$ in equation \ref{eqn:5bitSMPO_c0}.
% After the fifth clock cycle ends, this signal can be detected on wire $r_0$. It shows that 
% the result of $c_0$ is computed after 5 clock cycles and given on $r_0$.
% 
% If we track $R_1\to R_2\to R_3 \to \cdots \to R_0$, we can obtain $c_1$ respectively.
% Thus we conclude that Agnew's SMPO functions the same with Massey-Omura multiplier.
% \end{Example}
% 
% The design of Agnew's SMPO guarantees that there is only one AND gate in each $R_i$ block.
% For ONB, adopting Agnew's SMPO will reduce the number of AND gates from $2k-1$ to $k$.
% 
% \subsection{Multiplier not based on $\lambda$-Matrix}
% Both Massey-Omura multiplier and Agnew's SMPO rely on the implementation of $\lambda$-Matrix,
% which means that they will be identical if unrolled to full combinational circuits. 
% After Agnew's work of parallelization, researchers proposed more designs of SMPO, 
% some of them jump out of the circle and are independent from $\lambda$-Matrix.

>>>>

\section{Motivation}
\label{sec:normal_motiv}
From the preliminaries about FSMs in Section \ref{sec:reach}, we learn that the
Moore machine does not rely on inputs for state transitions. 
As depicted in Figure \ref{fig:Moore}(a), a typical Moore machine implementation
consists of combinational logic component and register files, where
$r_0,\dots,r_k$ are present state (PS) variables 
standing for state inputs (SI), and $r_0',\dots,r_k'$ are next state (NS) variables standing for
state outputs (SO). Figure \ref{fig:Moore}(b) shows the state transition graph (STG) of 
a Moore machine with $k+1$ distinct states. We notice that it forms a simple chain,
with $k$ consecutive transitions the machine reaches final state $R_k$.

@@

In practice,
some arithmetic components are designed in sequential circuits similar to the structure in 
Figure \ref{fig:Moore}(a). Initially the operands are loaded into the registers, 
then the closed circuit executes without taking any additional information from outside,
and store the results in registers after $k$ clock cycles. Its behavior can be described using
STG in Figure \ref{fig:Moore}(b): state $R$ denotes the bits stored in registers. Concretely, $R_{init}$ is the initial
state (usually reset to all zeros), $R_1$ to $R_{k-1}$ are intermediate results stored as SO of current state and SI
for next state, and $R_k$ (or $R_{final}$) is the final result given by arithmetic circuits (and equals to the answer
to arithmetic function when circuit is working functional correctly).
This kind of design results in 
reusing a smaller combinational logic component such that the area cost is greatly optimized.
An example of these designs is the sequential GF arithmetic multipliers described in Chapter \ref{ch:prelim_GF}.
However, it also brings difficulties in verifying the the circuit functions.

@@

\section{Formal Verification of Normal Basis Multiplier using Gr\"obner Basis}
The gate-level design of a NB multiplier can be generated using appoaches introduced in 
Section \ref{sec:nbdesign}. The gate-level netlist is ready to be verified using 
an approach similar to that in Chapter \ref{ch:reacha}. 
First we introduce the sketch of our approach using abstraction term order (ATO)
mentioned previously in Section \ref{sec:abstraction}, then refine our approach using 
the concept ``RATO", which is previously used in Section \ref{sec:improve}.
% The following part borrows contents
% from my own conference paper \cite{myDATE}.
\subsection{Implicit Unrolling based on Abstraction with ATO}
% fig:comb_RAB

The concept of abstraction was discussed in Section \ref{sec:abstraction}. If we use an elimination term order with
$intermediate~variables~>~R~>~A,B$ for the circuit in Figure \ref{fig:comb_RAB} to compute Gr\"obner basis, 
the function of the combinational logic component can be abstracted as 
$$R = \Func(A,B)$$
To verify the functional correctness of a combinational NB multiplier (e.g. Mastrovito multiplier
or Montgomery multiplier), the function given by abstraction will be computed as
$$R=A\cdot B$$
While in the sequential case, the function of combinational logic only fulfills a part of the multiplication.
For example, in the RH-SMPO introduced in Section \ref{sec:nbdesign}, 
the combinational logic actually implements  $F_{k-1}$ in Equation \ref{eqn:aux},
while computing all of $F_i$ requires $k$-cycle unrolling of the circuit.
Nevertheless, the abstraction still provides a word-level 
representation which can be more efficient for unrolling than bit-level expressions.
In other words, with the assistance of abstraction, we can execute implicit unrolling instead of 
explicit unrolling and avoid bit-blasting problem.

For 2-input sequential NB multipliers, abstraction is utilized to implement following algorithm:
%\vspace{-0.1in}
\IncMargin{1em}
\begin{algorithm}[hbt]
\SetAlgoNoLine
\LinesNumbered
 \KwIn{Circuit polynomial ideal $J$, vanishing ideal $J_0$, initial
   state ideal for output $R (=0)$, inputs $\mathcal{G}(A_{init}), \mathcal{H}(B_{init})$} 
\KwOut{Circuit function $R_{final}$ after $k$ cycles of $A_{init}, B_{init}$}
  $from_0(R,A,B) = \langle R, \mathcal{G}(A_{init}), \mathcal{H}(B_{init})\rangle$\;
  \tcc{$A, B$ are word-level variables, solutions to polynomial equations 
  $\mathcal{G},\mathcal{H}$ denote the initial values of preloaded operands $A_{init},B_{init}$}
  $i = 0$\;
  \Repeat{$i == k$}
  {
  	$i \gets i + 1$\;
	$G \gets$GB$( \langle J + J_0+ from_{i-1}(R,A,B) \rangle
    )$ with ATO\;
	$to_i(R',A',B')\gets G\cap \mathbb F_{2^k}[R',A',B',R,A,B]$\;
	\tcc{Using projections of varieties from abstraction theory, we get NS variables $R',A',B'$
	in terms of PS $A,B$}
	$from_i \gets to_i(\{R,A,B\}\setminus \{R',A',B'\})$\;
	\tcc{By replacing NS variables with PS variables we push it to next time-frame}
  }
\Return{$from_k(R_{final})$}
\caption {Abstraction via implicit unrolling for Sequential GF circuit
  verification}
\label{alg:modified}
\end{algorithm}
\DecMargin{1em}

Notations in the following figure as well as in next example comply with 
the notations in this algorithm.
